\documentclass[useAMS,usenatbib]{mn2e}
\usepackage{epsfig,rotate,graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{subfigure}
\usepackage{lscape}
\usepackage{bm}

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJL}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle_\phi}
\newcommand{\bu}{\bm{u}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\teta}{\tilde{\eta}}
\newcommand{\tcool}{t_\mathrm{cool}}
\DeclareMathOperator{\erf}{erf}

\def\la{\left\langle\rule{0pt}{3em}}
\def\ra{\right\rangle}

\usepackage{array,booktabs,tabularx}
\newcolumntype{R}{>{\centering\arraybackslash}X} % right justified tabularx columns


\title[Gaps in non-isothermal discs]{Gap formation and stability in 
  non-isothermal protoplanetary discs} 

\author[Les and Lin]{Robert Les
  $^1$\thanks{robert.les@mail.utoronto.ca} and Min-Kai Lin $^{1,2}$
  \thanks{ minkailin@email.arizona.edu} \\ 
  $^1$Canadian Institute for Theoretical Astrophysics,  
  60 St. George Street, Toronto, ON, M5S 3H8, Canada \\
  $^2$Department of Astronomy and Steward Observatory, University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA 
}

\begin{document}

\maketitle
\begin{abstract}
  Several observations of transition discs show lopsided
  dust-distributions. A potential explanation is the formation of a
  large-scale vortex acting as a dust-trap at the edge of a gap opened
  by a giant planet. Numerical models of gap-edge vortices have
  thus far employed locally isothermal discs, but the 
  theory of this vortex-forming or `Rossby wave' instability was
  originally developed for adiabatic discs.  
  We generalise the study of planetary gap stability to non-isothermal
  discs using customised numerical simulations of disc-planet
  systems where the planet opens an unstable gap. 
  We include in the energy equation a simple cooling function with
  cooling timescale $t_c=\beta\Omega_k^{-1}$, where $\Omega_k$ is
  the Keplerian frequency, and examine the effect of $\beta$ on the
  stability of gap edges and vortex lifetimes. We find increasing
  $\beta$ lowers the growth rate of non-axisymmetric perturbations, and the
  dominant azimuthal wavenumber $m$ decreases.  
  We find a quasi-steady state consisting of one  
  large-scale, over-dense vortex circulating the outer gap edge, typically  
  lasting $O(10^3)$ orbits. 
  Vortex lifetimes were found to generally increase with cooling times up to an
  optimal value, beyond which vortex lifetimes decrease. This
  non-monotonic dependence is qualitatively consistent with 
  recent studies using strictly isothermal discs that vary the disc aspect ratio. 
%  presented in Fu et al. (2014, ApJL, 788, L41). 
\end{abstract}

\begin{keywords}
  accretion, accretion discs, protoplanetary discs, hydrodynamics, instabilities,
  planet-disc interactions, methods: numerical 
\end{keywords}


\section{Introduction}\label{intro}
%{
%\bf 
%observational motivation: asymmetric dust distributions. 
%theoretical motivations: generalize gap stability to include energy equation. 
%}
The interaction between planets and protoplantary discs plays an
important role in the theory of planet formation and disc 
evolution. Disc-planet interaction may lead to the orbital migration
of protoplanets and modify the structure of
protoplanetary discs  \citep[see][for a recent review]{baruteau13}.  
%\citep[][see \cite{baruteau13} for a recent
%  review]{goldreich79, lin79, goldreich80,lin86}. 

A sufficiently massive planet can open a gap in a 
gaseous protoplanetary disc \citep{pap_lin84,bryden99,crida06,fung14}, 
while low mass planets may also open gaps if the disc viscosity is
small enough \citep{li09,dong11,duffell13}. Support for such disc-planet
interaction have begun to emerge in observations of circumstellar
discs that reveal annular gaps
\citep[e.g.][]{quanz13a,debes13,osorio14}, with possible evidence of
companions within them \citep[e.g.][]{quanz13b,reggiani14}. 

%gaps can be unstable - vortices 

A recent theoretical development in the study of planetary gaps is
their stability. When the disc viscosity is low and/or the planet mass
is large, the presence of potential vorticity (PV, the ratio of
vorticity to surface density) extrema can render planetary gaps
dynamically unstable due to what is now referred to as the `Rossby
wave instability' \citep[RWI,][]{lovelace99,li00}. This 
eventually leads to vortex formation 
\citep{li01,koller03,li05,valborro07}, which can significantly affect
orbital migration of the planet  \citep{ou07,li09,yu10,lin10}. 

Vortex formation at gap edges may also have observable 
consequences. Because disc vortices represent pressure maxima, they are
able to collect dust particles 
\citep{barge95,inaba06,lyra13}. Dust-trapping at gap-edge vortices
have thus been suggested to explain asymmetric dust
distributions observed in several transition discs
\citep[e.g][]{marel13,isella13,perez14}. 

However, studies of Rossby vortices at planetary gap-edges have 
adopted locally isothermal discs, %cite lin11, zhu14, lyra
where the disc temperature is a fixed function of
position only \citep[e.g.][]{lyra08,lin11a,zhu14,fu14}. 
%isothermal valid for outer disc  
On the other hand, the theory of the RWI was in fact
developed for adiabatic discs \citep{li00}, which permits
heating. %cite 
In adiabatic discs, the relevant quantity for stability
becomes a generalisation of the PV that include entropy variations
\citep{lovelace99}.   


Gap-opening is associated with planet-induced spiral shocks. In an
isothermal disc, PV-generation across these isothermal shocks leads to
the RWI \citep{koller03,li05,valborro07,lin10}.    
However, if cooling is inefficient and the shock is non-isothermal,
then shock-heating may affect gap stability, since the
relevant quantity is an entropy-modified PV (described below), and
there is entropy-generation across the shocks. 

For example, previous
linear simulations of the RWI found  
that increasing the sound-speed favours instability \citep{li00,lin13}.  
In the context of
planetary gaps, however, the increased temperature may also act to
stabilise the disc by making gap-opening more difficult. It is 
therefore of theoretical interest to clarify the effect of heating and
cooling on the stability of planetary gaps. 

In this work, we extend the study of planetary gap stability against
vortex formation to non-isothermal discs. We include in the fluid energy
equation an one-parameter cooling prescription that allows us to probe
disc thermodynamics ranging from nearly isothermal to nearly
adiabatic.      

This paper is organised as follows. In \S\ref{model} we describe the
equations governing the disc-planet system and initial conditions. Our
numerical approach, including diagnositic measures, are given in
\S\ref{method}. We present results from two sets of numerical
experiments. In \S\ref{linear1} we use disc-planet interaction to set
up discs with gaps, but study their stability without further
influence from the planet. %stability through background state 
We then perform long-term disc-planet simulations to examine the
lifetime of gap-edge vortices in \S\ref{nonlinear},  
as a function of the imposed cooling rate. We conclude and summarize
in \S\ref{summary} with a discussion of important caveats. 



\input{model}
\input{results}
%\input{discussion}
\input{summary}

%\section*{Acknowledgments}

\bibliographystyle{mn2e}
\bibliography{ref}

\appendix
%\input{appendix}
%\input{appendix2}

\end{document}
